### 2014-11-23 doc/thesis/system.tex
Starting to write about the refinement of table types

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\vec{l} = el \; | \;
m \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
\mathbf{return} \; el \; | \;
e(el)_{s} \; | \;
e{:}n(el)_{s}\\
e ::= & \;\; \mathbf{nil} \; | \;
\mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string} \; | \;
{...}_{e} \; | \;
f \; | \;
\{ \} \; | \;
\{ \; cl \; \} \\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
e(el)_{e} \; | \;
e{:}n(el)_{e} \; | \;
n \; | \;
e_{1}[e_{2}] \; | \;
{<}t{>} \; n\\
l ::= & \;\; n \; | \;
e_{1}[e_{2}] \; | \;
n[e] \; {<}t{>}\\
el ::= & \;\; \mathbf{nothing} \; | \;
\vec{e} \; | \;
me \; | \;
\vec{e}, me\\
cl ::= & \;\; \vec{c} \; | \;
\vec{e} \; | \;
me \; | \;
\vec{c}, \vec{e} \; | \;
\vec{c}, me \; | \;
\vec{c}, \vec{e}, me \; | \;
\vec{e}, me\\
me ::= & \;\; e(el)_{m} \; | \;
e{:}n(el)_{m} \; | \;
{...}_{m}\\
m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; ({...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t},{...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
f ::= & \;\; \mathbf{fun} \; (){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; ({...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; (\vec{n{:}t}){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
c ::= & \;\; [e_{1}] = e_{2}\\
n ::= & \;\; {\it name}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}

@@

@@

Multiple assignments and multiple return values can also appear in
function applications, as the following example shows:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; m:\Void \rightarrow \Integer \times \String =$} \\
& \multicolumn{3}{l}{$\mathbf{fun} \; ():\Integer \times \String$} \\
& & \multicolumn{2}{l}{$\mathbf{skip} \;;\; \mathbf{return} \; 2, ``foo"$} \\
\multicolumn{4}{l}{$\mathbf{in}$} \\
& \multicolumn{3}{l}{$\mathbf{local} \; s:\Integer \times \Integer \rightarrow \Integer =$} \\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer, y:\Integer):\Integer$} \\
& & & \multicolumn{1}{l}{$\mathbf{skip} \;;\; \mathbf{return} \; x + y$} \\
& \multicolumn{3}{l}{$\mathbf{in} \; s(m()_{e}, m()_{m})_{s}$}
\end{tabular}
\end{center}

@@

Refinement of table types (first example).
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; t:\{\}_{open} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$t[``f"] \; {<}\String{>} = ``Lou";$}\\
& \multicolumn{1}{l}{$t[``l"] \; {<}\String{>} = ``Reed";$}\\
\end{tabular}
\end{center}

>>>>

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
ll = el_{1} \; | \;
m \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n{:}t} = el_{0} \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
\mathbf{return} \; el\\
& | \; e(el)_{s} \; | \;
e{:}n(el)_{s}\\
e ::= & \;\; \mathbf{nil} \; | \;
x \; | \;
{...}_{e} \; | \;
n \; | \;
e_{1}[e_{2}] \; | \;
{<}t{>} \; n \; | \;
f \; | \;
\{ \} \; | \;
\{ \; cl \; \} \\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
e(el)_{e} \; | \;
e{:}n(el)_{e}\\
x ::= & \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string}\\
l ::= & \;\; n \; | \;
e_{1}[e_{2}] \; | \;
n[e] \; {<}t{>}\\
ll ::= & \;\; \vec{l}\\
el_{0} ::= & \;\; \mathbf{nothing} \; | \;
el_{1}\\
el_{1} ::= & \;\; \vec{e} \; | \;
me \; | \;
\vec{e}, me\\
el ::= & \;\; el_{0}\\
cl ::= & \;\; \vec{c} \; | \;
{...}_{m} \; | \;
\vec{c}, {...}_{m}\\
me ::= & \;\; e(el)_{m} \; | \;
e{:}n(el)_{m} \; | \;
{...}_{m}\\
m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; ({...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t},{...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
f ::= & \;\; \mathbf{fun} \; (){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; ({...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; (\vec{n{:}t}){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
c ::= & \;\; [e_{1}] = e_{2}\\
n ::= & \;\; {\it name}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}

@@

The most unusual feature of our type system is the refinement of
table types, which is given by the following rule:
\[
\begin{array}{c}
\mylabel{T-REFINE}\\
\dfrac{\env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique} \;\;\;
       \env_{1} \vdash x:t_{1}, \env_{2} \;\;\;
       \not \exists i \in 1..n \; t_{1} \lesssim k_{i}}
      {\env_{1} \vdash n[x] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t\}_{open|unique}]}
\end{array}
\]

The rule \textsc{T-REFINE} allow us to add new keys to \emph{open}
or \emph{unique} table types.
We restricted the keys to literals only because the purpose of
this rule is to make it easier to add new fields to table types
that represent records.
We also do not allow changing the type of a field that is already
in the table because this could lead to loosen table types that
have no static type checking. 
For instance, Typed Lua allows us to write the following code:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; t:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$t[``x"] \; {<}\String{>} = ``foo";$}\\
& \multicolumn{1}{l}{$t[``y"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}
but it does not allow us to write the following code:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; t:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$t[``x"] \; {<}\String{>} = ``foo";$}\\
& \multicolumn{1}{l}{$t[``x"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}

@@

Multiple assignments and multiple return values can also appear in
function applications, as the following example shows:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; m:\Void \rightarrow \Integer \times \String =$} \\
& \multicolumn{3}{l}{$\mathbf{fun} \; ():\Integer \times \String$} \\
& & \multicolumn{2}{l}{$\mathbf{return} \; 2, ``foo"$} \\
\multicolumn{4}{l}{$\mathbf{in}$} \\
& \multicolumn{3}{l}{$\mathbf{local} \; s:\Integer \times \Integer \rightarrow \Integer =$} \\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer, y:\Integer):\Integer$} \\
& & & \multicolumn{1}{l}{$\mathbf{return} \; x + y$} \\
& \multicolumn{3}{l}{$\mathbf{in} \; s(m()_{e}, m()_{m})_{s}$}
\end{tabular}
\end{center}

@@
